---
title: "Data Visualisation with ggplot2 - Project"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# AfterWork Data Science: Data Visualisation with Python Project

## 1. Defining the Question

### a) Specifying the Data Analysis Question
Data science analysis to provide Hilton Hotel International with recommendation on strategies that would lead to revenue growth.

### b) Defining the Metric for Success
The solutions to the following questions will help us answer our research question:

* When is the best time of year to book a hotel room?
* When is the optimal length of stay in order to get the best daily rate?
* How will you know if a hotel was likely to receive a disproportionately high number of special requests?

### c) Understanding the context
The data is originally from the article Hotel Booking Demand Datasets, by Nuno Antonio, Ana Almeida, and Luis Nunes for Data in Brief, Volume 22, February 2019.

### d) Recording the Experimental Design
How many bookings were canceled?
● What was the booking ratio between resort hotels and city hotels?
● What was the percentage of booking for each year?
● Which was the busiest month for hotels?
● Most guests came from which country?
● How long do most people stay in hotels?
● Which was the most booked accommodation type (Single, Couple, Family)?

### e) Data Relevance
How data was relevant to our question.


## 2. Reading the Data

```{r}
# Load the data below
# --- 
# Dataset url = https://bit.ly/2WvQbhJ
# --- 
# YOUR CODE GOES BELOW
# 
hotel_booking_df <- read.csv("hotel_bookings.csv", header = TRUE, na.strings=c("NULL",NA))
hotel_booking_df
```

```{r}
# Checking the first 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 
head(hotel_booking_df, 5)
```

```{r}
# Checking the last 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 
tail(hotel_booking_df, 5)
```

```{r}
# Sample 10 rows of data
# ---
# YOUR CODE GOES BELOW
# 
sample_n(hotel_booking_df, 10)
```

```{r}
# Checking number of rows and columns
# ---
# YOUR CODE GOES BELOW
#  
dim(hotel_booking_df)
```

```{r}
# Checking datatypes
# ---
# YOUR CODE GOES BELOW
# 
str(hotel_booking_df)
```

```{r}

```

Record your general observations below:

Observation 1:It was observed that there were columns with null values i.e agent and company.
Observation 2


## 3. External Data Source Validation
The data is originally from the article Hotel Booking Demand Datasets, by Nuno Antonio, Ana Almeida, and Luis Nunes for Data in Brief, Volume 22, February 2019.


## 4. Data Preparation
### Performing Data Cleaning
```{r}
# Checking datatypes and missing entries of all the variables
# ---
# YOUR CODE GOES BELOW
# 
str(hotel_booking_df)
sum(is.na(hotel_booking_df))
```

We observe the following from our dataset:

Observation 1
Observation 2: There are 129425 missing values in our data set

```{r}
# Checking how many duplicate rows are there in the data
# ---
# YOUR CODE GOES BELOW
# 
#
sum(duplicated(hotel_booking_df))
```

I got 31994 duplicated entries .

```{r}
# Checking if any of the columns are all null
# ---
# YOUR CODE GOES BELOW
# 
which(colSums(is.na(hotel_booking_df))  == ncol(hotel_booking_df))
```


Observation 1: I could not find any columns in which all the data is null

```{r}
# Checking if any of the rows are all null
# ---
# YOUR CODE GOES BELOW

which(rowSums(is.na(hotel_booking_df)) == nrow(hotel_booking_df))
```


Observation 1:  I could not find any row in which all the data is null
Observation 2

```{r}
# Checking the correlation of the features through the use of 
# visualizations the correlation using heatmap
# ---
# YOUR CODE GOES BELOW
# 

sample(hotel_booking_df,10)

# Select the desired columns
hotel_booking_df2 <- hotel_booking_df %>%
  na.omit() %>%
  select(is_canceled, stays_in_weekend_nights, stays_in_week_nights, adults, children, babies, is_repeated_guest, previous_cancellations, previous_bookings_not_canceled, days_in_waiting_list, total_of_special_requests)
hotel_booking_df2

# Create a correlation matrix
corr_matrix <- cor(hotel_booking_df2, method="s")
head(corr_matrix)

# Create a table with long form
corr_df <- melt(corr_matrix)
corr_df

# Plot the heatmap
ggplot(corr_df, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "black") +
  geom_text(
    aes(label = round(value, 2)),
    color = "white"
  ) +
  coord_fixed() +
  labs(
    fill="Pearson Correlation"
  ) +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white",
    limit = c(-1,1)
  ) +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )

```

We observe the following from our dataset:

Observation 1
Observation 2

```{r}
# Dropping company column because it has alot of missing values 
# and we won't need to answer any of our questions
# ---
# YOUR CODE GOES BELOW
# 
hotel_booking_df = select(hotel_booking_df, -company)
head(hotel_booking_df)
```

From the data variable description we see that the Distribution Channel category that tells us about Booking distribution.

The term “TA” means “Travel Agents”
The term “TO” means “Tour Operators”
This allows us to fill the missing values in the agents column with TO

```{r}
# We replace the mising values i.e. for TO
# ---
# YOUR GOES BELOW
#

hotel_booking_df$agent[is.na(hotel_booking_df$agent)] <- "TO"
hotel_booking_df
```

```{r}
# We drop rows where there is no adult, baby and child as 
# these records won't help us.
# ---
# YOUR GOES BELOW
#

hotel_booking_df <- subset(hotel_booking_df, adults!=0 | children!=0 | babies!=0)
hotel_booking_df
```

```{r}
# We replace missing children values with rounded mean value
# ---
# Hint i.e. use round()
# ---
# YOUR GOES BELOW
#
hotel_booking_df$children[is.na(hotel_booking_df$children)] <- round(mean(hotel_booking_df$children, na.rm = TRUE), 0)
hotel_booking_df
```

```{r}
# Checking for missing values in the dataframe
# ---
# YOUR GOES BELOW
#
sum(is.na(hotel_booking_df))
```

```{r}
# Converting the datatypes of the following columns from float to integer
# i.e. children, company, agent
# ---
# YOUR GOES BELOW
# 
# hotel_booking_df$children <- lapply(hotel_booking_df$children, as.numeric)
as.integer(hotel_booking_df$children)
str(hotel_booking_df)
```


## 5. Solution Implementation

### 5.a) Questions

```{r}
# 1. How many bookings were cancelled?
# ---
# Visualisation: Barplot

hotel_booking_df %>%
  group_by(reservation_status) %>%
  summarise(is_canceled = sum(is_canceled))

ggplot(hotel_booking_df, aes(x= reservation_status, y= is_canceled))+
  geom_col(stat = "identity", width =  0.5, fill= "blue") +
  labs(
    title = "Bar graph of Visitors Reservation Status",
    x= "Booking Status",
    y= "Booking Number"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Ariel Black", color="red")
  )
  
```


```{r}
# 2. What was the booking ratio between resort hotel and city hotel?
# ---
# pie chart showcase % booking ratio between resort hotel and city hotel very well

resort_city_hotel <- hotel_booking_df %>%
  count(hotel)
resort_city_hotel

ggplot(resort_city_hotel, aes(x="", y=n, fill=hotel)) +
  geom_bar(
    stat="identity", width=0.5, color = "yellow"
  ) +
  coord_polar("y") +
  geom_text(
    aes(label = paste0(round(n / sum(n) * 100, 2), "%")), position = position_stack(vjust = 0.5)) +
  scale_fill_brewer(palette="Set2")+
  labs(
    fill = "Hotels names",
    title = "showcase % booking ratio between resort hotel and city hotel"
  )+
  theme_void()
```


```{r}
# 3. What was the percentage of booking for each year?
# ---
# pie chart showcase % booking for each year very well

year_booking <- hotel_booking_df %>%
  count(arrival_date_year)
year_booking

ggplot(data=year_booking, aes(x="", y=n, fill=factor(arrival_date_year))) +
  geom_bar(
    stat="identity", width=0.5, color = "yellow"
  ) +
  coord_polar("y") +
    geom_text(
    aes(label = paste0(round(n / sum(n) * 100, 2), "%")), position = position_stack(vjust = 0.5)
  ) +
  scale_fill_brewer(palette="Set1") +
  labs(
    fill = "Years of booking",
    title = "percentage booking for each year") +
  theme_void()
```

```{r}
# 4. Which were the most busiest months for hotels?
# ---
# Bar chart was the most appropriate here

busy_months <- hotel_booking_df %>%
  count(arrival_date_month)
busy_months

ggplot(busy_months, aes(x= arrival_date_month, y= n))+
  geom_col(stat = "identity", width =  0.8, fill= "magenta") +
  labs(
    title = "Hotel Bookings per Month",
    x= "Month",
    y= "Number of bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Arial", color="blue")
  )
```


```{r}
# 5. From which top 3 countries did most guests come from?
# --- 
# YOUR code GOES BELOW
#

#hotel_booking_df

countries <- hotel_booking_df %>%
  group_by(country) %>%
  summarise(guests = sum(adults)+sum(children)+sum(babies)) %>%
  # arrange(desc(guests)) %>%
  top_n(3)
countries

ggplot(countries, aes(x= country, y= guests))+
  geom_col(stat = "identity", width =  0.6, fill= "green") +
  labs(
    title = "Top 3 Countries per Guests",
    x= "Countries",
    y= "Number of Guests"
  ) +
  theme(
    plot.title = element_text(hjust = 0.4, size=16, family="Arial", color="black")
  )
```

```{r}
# 6.a) How long do most people stay in hotels?
#   b) By city and resort? Separate the data by hotel
# ---
# 

stay_length <- hotel_booking_df %>%
  group_by(hotel) %>%
  summarise(total_stay_days = mean(stays_in_weekend_nights)+mean(stays_in_week_nights))
stay_length

ggplot(data = stay_length, aes(x= hotel, y= total_stay_days))+
  geom_col(stat = "identity", width =  0.4, fill= "blue") +
  labs(
    title = "Days of Stay in Hotel",
    x= "Type of Hotel",
    y= "Duration of stay in days"
  ) +
  theme(
    plot.title = element_text(hjust = 0, size=16, family="Arial", color="blue")
  )+
  theme_classic()
```


```{r}
# 7. Which was the most booked accommodation type (Single, Couple, Family)?
# ---
#
accommodation <- hotel_booking_df %>%
  count(reserved_room_type)
accommodation

ggplot(data = accommodation, aes(x= reserved_room_type, y= n))+
  geom_col(stat = "identity", width =  0.6, fill= "pink") +
  labs(
    title = "Bookings per Accomodation Type",
    x= "Room Type",
    y= "Number of Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Arial", color="blue")
  )
```

### 5.b) Recommendations
Recommendation:
1.The company should find reasons why a good number of visitors are cancelling the their bookings and turn them to promoters.
2.The company should invest more on city facilities/comfort since its attracts 66.41% of total customers while also paying attention to Resort Hotel
3.The company should increase its budget on advertisement/marketing/facilities upgrade yoy to ensure visitors booking keeps growing.from data <2015,2016,2017> it showing reduction on hotel bookings.
4.for the months with over 10000 bookings, the company should invest on its preparedness to ensure customer experince is uncompromised and hence boost NPS
5.Since portugal guests contribute slightly less than 50% of guests, they should be considered while preparing for their reception i.e provide their cultural meals and customize their environment.
6.Type A of rooms attracts over 50% of guests, therefore its ammenities should be upgraded to world class to ensure customer experience excellence.





## 6. Challenging your Solution
In this step, we review our solution and implement approaches that could potentially provide a better outcome. In our case, we could propose the following question that wasn't answered in our solution because it couldn't have greatly contributed to our recommendation.

```{r}
# When should hotels provide special offers?
# ---
# YOUR  answer GOES BELOW:
#special offers should be given any time there is <7500 number of bookings to encourage more visitors.this will be during the month of January, November and December.

# 

```


Our observations:
-

How does this observation tie to our solution?
Anwer: encourages more visitors to visisit during low seasons.this will be during the month of January, November and December

## 7. Follow up questions
During this step, you rethink and propose other ways that you can improve your solution.

a). Did we have the right data?Yes.
b). Do we need other data to answer our question? No, the one we have was sufficient.
c). Did we have the right question? yes we did.


